Modeling Intervention Strategies for United States TB Control

### 2014-02-15 in_progress/presentations/paper/draft.tex
Starting agent-based methods section

\section{Methods}
\subsection{Basic Structure}
Both the Hill Model (Basic Hill Model) and our extended Model (Extended Hill
Model) were implemented in \texttt{R} as a system of differential equations,
which were solved via the $\texttt{lsoda}$ routine. The systems of differential
equations used to capture basic disease dynamics in the Basic Hill Model are
shown in Figure~\ref{fig:hillEquations}. In the Extended Hill Model, more
differential equations were implemented to add additional tracking and economic
modeling. These equations are detailed in the\\
{\huge\color{red} Should appendix be capitalized? What about Figure?}\\
appendix,
section~\ref{sec:extendedHillEqs}.  Historical data from 2000-2008 was used to
initialize the model, which was then run up to the year 2100.

@@

The vector variables $S_0, F_0, L_0, I_0, J_0$ contain the number of US-born
individuals in the S, F, L, I, J compartments respectively, whereas $S_1, F_1,
L_1, I_1, J_1$ contain the number of foreign-born individuals. $N_0$ and $N_1$
are the total populations of US-born and foreign-born individuals.  The
constants $\rho$ and $\alpha$ are birth rates, while $\mu_{i}$, and $\mu_{d}$
are death rates. A complete list and descriptions of all constants used in the
model can be found in the Appendix, Section~\ref{subsec:hillConstants}.

\subsection{Additional Tracking Capabilities}
In order to refine the tracking capabilities of the
Hill Model, the original differential equations used to describe TB spread were
separated into their component parts and each section was tracked separately.
These components were made into compartments, tracked by differential equations
detailed in the appendix. These equations allowed the model to track the total
number of TB cases, TB deaths, and  natural deaths. Further, the model tracks,
the sourced cost on the US HCS. Progressions into
active TB due to activations of LTBI or exogenous infection were tracked,
allowing for sourced incidence data to be generated. The entering cases of LTBI
(acute or chronic) were also tracked. In the case of intervention testing, the
number of cured and untreated cases of entering LTBI were both tracked. Cured
cases of LTBI entering, TB deaths, total TB cases, and total cost were also
tracked discounted at 3\% annually. This discounting was converted to a
continuous differential equation for use within the model. In general, incidence
data is calculated in the same way in the Extended Hill Model as it was in the Basic
Hill Model.

In addition, estimations of the basic reproductive number of FB or USB active,
infectious TB cases were made from a theoretical and an experimental
perspective. Experimental data were calculated by reducing the initial
population of FB or USB active, infectious TB cases by one and allowing the
model to progress otherwise as normal. The decreased number of total TB cases
seen details how many infections can be thought to be due to one infected
individual in the given population.

From an experimental perspective, the spread of TB was thought to be a geometric
series. If one infectious individual infects $x$ people annually, over the
course of $N$ years the total number of infections caused by this individual can
be approximated by the geometric series of $N$ terms with rate $x$ and initial
term $1$. In this case, $N = 100$, and the ratios for FB infectious individuals
and USB infectious individuals were obtained from RATIO CITATION COLIN. 
\subsection{Economic Modeling}
When tracking the economic implications of current US TB load and various
intervention strategies, active TB costs were given by \$14,000. This cost is
the weighted mean of the costs of cases requiring hospitalization (happening
49\% of the time) and cases not requiring hospitalization (51\%). These data
were found in DYLAN COST STUDY CITATION. LTBI treatment costs were given by
\$468.00. These data were based on the cost of a successfull treatment, given by
DYLAN COST STUDY CITATION, as well as adherence and efficacy data found in
ADHERENCE CITATION.

From these estimated cost per treatment, total costs were obtained by assuming
that every patient with active TB in the US is treated at full price, though the
treatment may fail. As individuals with active
TB are charged upon entry to the active TB compartment, the model can also
accurately estimate sourced US HCS cost data. Cost data were separately tracked
due to incoming active TB costs stemming from activation of LTBI vs exogenous
reinfection. These data were modelled as additional compartments and also solved
by $\texttt{lsoda}$. For LTBI treatment cost, treatment was
charged upon leaving the LTBI compartment due to the cumulative self-cure and
treatment rate given in the Hill Model. The fraction of these indviduals who
leave this compartment due to self cure was assumed to be zero. Given the
uncertainty in measurments of LTBI treatment cost, extensive sensitivity
analysis was performed on this parameter relative to final cost outcomes. In
addition to US HCS cost, the system was implemented so as to also track
projected intervention implementation cost, given user-inputted parameters
relating to various possible intervention cost strategies. The extended Hill
Model was used to track the effect of many interventions tested by the Hill
Model. In particular, the intervention strategy of curing entering cases of LTBI
proved very promising and elimination year, final cost, and cost per case
averted were tracked for various levels of entering LTBI cure rate. The
interventions were implemented so as to take effect during the year 2013 and run
to 2100. The numerical DE solver $\texttt{lsoda}$ used was run with a time step
of $0.8$. Sensitivity analysis was performed on this parameter and reducing the
time step was shown to have minimal effect on final size or cost values.

>>>>

\section{Methods}
\subsection{Basic Structure}
Both the Hill Model (Basic Hill Model) and our extended Model (Extended Hill
Model) were implemented in \texttt{R} as a system of differential equations,
which were solved via the $\texttt{lsoda}$ routine. The systems of differential
equations used to capture basic disease dynamics in the Basic Hill Model are
shown in Figure~\ref{fig:hillEquations}.  In the Extended Hill Model, each flow
rate equation from the Basic Hill Model was separated into its component part
and these were tracked separately. In other words, rather than tracking net flow
into various compartments, the Extended Hill Model tracks flow between all pairs
of compartments separately. Furthermore, these equations were each also supplemented to
also estimate their corresponding US HCS costs.  These equations are detailed in
the\\
{\huge\color{red} Should appendix be capitalized? What about Figure?}\\
appendix,
section~\ref{sec:extendedHillEqs}.  Historical data from 2000-2008 was used to
initialize the model, which was then run up to the year 2100.

@@

The vector variables $S_0, F_0, L_0, I_0, J_0$ contain the number of US-born
individuals in the S, F, L, I, J compartments respectively, whereas $S_1, F_1,
L_1, I_1, J_1$ contain the number of foreign-born individuals. $N_0$ and $N_1$
are the total populations of US-born and foreign-born individuals.  The
constants $\rho$ and $\alpha$ are birth rates, while $\mu_i$, and $\mu_d$ are
death rates. A complete list and descriptions of all constants used in the model
can be found in the Appendix, Section~\ref{subsec:hillConstants}.

\subsection{Additional Tracking Capabilities}
In order to refine the tracking capabilities of the Hill Model, the original
differential equations used to describe TB spread were separated into their
component parts and each section was tracked separately.
These components were made into compartments, tracked by differential equations
detailed in the appendix. Progressions into active TB due to activations of LTBI
or exogenous infection were tracked, allowing for sourced incidence data to be
generated. These equations allowed the model to track the sourced number of TB
cases, TB deaths, and natural deaths. Furthermore, the model also tracks the
sourced cost on the US HCS. The immigrating cases of LTBI (acute or chronic)
were also tracked. In the case of intervention testing, the number of cured and
untreated cases of entering LTBI were both tracked. Cured cases of LTBI
entering, TB deaths, total TB cases, and total cost were also tracked discounted
at 3\% annually. This discounting was converted to a continuous differential
equation for use within the model. In general, incidence data is calculated in
the same way in the Extended Hill Model as it was in the Basic Hill Model.

In addition, estimations of the basic reproductive number of FB or USB active,
infectious TB cases were made from a theoretical and an experimental
perspective. Experimental data were calculated by reducing the initial
FB or USB infectious TB populations by one and allowing the
model to progress otherwise as normal. The decreased number of total TB cases
seen details how many infections can be thought to be due to one infected
individual in the given population. In the agent-based model, this \emph{one}
was a true model agent whereas in the compartmental model, this decrease of
\emph{one} was a proportional decrease based on the scaled population level.

From a theoretical perspective, the spread of TB was thought to be a geometric
series. If one infectious individual infects $x$ people annually, over the
course of $N$ years the total number of infections caused by this individual can
be approximated by the geometric series of $N$ terms with rate $x$ and initial
term $1$. In this case, $N = 100$, and the ratios for FB infectious individuals
and USB infectious individuals were obtained from <+RATIO CITATION COLIN+>
\subsection{Economic Modeling}
In order to track the US HCS economic load given the TB rates, we made the
following treatment cost assumptions. A single case of active TB was assumed to
yield a \$14,000 US HCS cost, charged immediately upon disease contraction. This
cost is the weighted mean of the costs of cases requiring hospitalization
(happening 49\% of the time) and cases not requiring hospitalization (51\%).
These cost estimates were found in <+DYLAN COST STUDY CITATION+>. LTBI treatment
costs were given by \$468.00 US HCS cost, charged immediately upon successful
treatment. These estimates were calculated based on the cost of a successfull
treatment, and the typical adherence and efficacy of said treatment <+DYLAN COST
STUDY CITATION, ADHERENCE CITATION+>. It is important to note that the charging
assumptions are different for treatment of active TB as opposed to treatment of
latent TB. In particular, cases of active TB imply a cost immediately upon
disease contraction, whereas latent TB costs are charged only upon successful
treatment. The motivation for this distinction is that active TB treatment is
mandated by the US for all known cases and, further, most treatment costs are
incurred at the beginning of the treatment cycle. As such, charging for every
known case, immediately at contraction is a well motivated assumption. On the
other hand, latent TB is only rarely treated in discovered cases and the costs
are more evenly distributed. Furthermore, the choice to charge only upon
successful treatment yields a more conservative estimate of the success of any
intervention analyzed.

Cost data were separately tracked due to incoming active TB costs stemming from
activation of LTBI vs exogenous reinfection. These data were modelled as
additional compartments and also solved by $\texttt{lsoda}$.  Given the
uncertainty in measurments of LTBI treatment cost, extensive sensitivity
analysis was performed on this parameter relative to final cost outcomes. In
addition to US HCS cost, the system was implemented so as to also track
projected intervention implementation cost, given user-inputted parameters
relating to various possible intervention cost strategies. The extended Hill
Model was used to track the effect of many interventions tested by the Hill
Model. In particular, the intervention strategy of curing entering cases of LTBI
proved very promising and elimination year, final cost, and cost per case
averted were tracked for various levels of entering LTBI cure rate. The
interventions were implemented so as to take effect during the year 2013 and run
to 2100. The numerical DE solver $\texttt{lsoda}$ used was run with a time step
of $0.8$. Sensitivity analysis was performed on this parameter and reducing the
time step was shown to have minimal effect on final size or cost values.
